# Website Setup

This documents how I set up the `animint2` documentation website with `pkgdown`. It is also the code for setting up the website.

See [this webpage for more information on using `pkgdown`](https://pkgdown.r-lib.org/articles/pkgdown.html).


## Build the Site

First, we use `usethis` to get `animint2` ready to become a website. The code only needs to be run once, and I've run it already:

```{r, eval=FALSE}
usethis::use_pkgdown()
usethis::use_pkgdown_github_pages()
```

This does three things:

- Creates the `_pkgdown.yml` file, which we'll need to edit later.
- Adds the following regular expressions to the `.Rbuildignore` file: `^_pkgdown\\.yml$`, `^docs$`, and `^pkgdown$`. When the `animint2` package is built, it'll ignore the website stuff.
- Adds the `docs/` directory to the `.gitignore` file.

The last bullet point may be surprising. Why are we telling git to ignore our nicely-rendered website? Because we're using Github, and there's another function that will make sure our website appears on Github Pages.

Next, we need to convert `README.org` into `README.md`. `pkgdown` doesn't understand Org and requires a README to generate the website's homepage. I used `pandoc`, but I guess you could also convert it by hand.

```
pandoc -s README.org -o README.md
```

Then we generate the website:

```{r}
pkgdown::build_site()
```

Finally, we 


## Making Changes to the Website

The website works, but it's disordered. The site is mostly a mix of the Github README and the official `animint2` documentation on CRAN. It could be better.












