# Website Setup

This documents how I set up the `animint2` documentation website with `pkgdown`. It is also the code for setting up the website.


## Build the Site

First, we use `usethis` to get `animint2` ready to become a website. We'll also need `gitcreds`, since we want The code only needs to be run once, and I've run it already. But in case you need to run it again, let's go through what you need to do. First, we'll generate a GitHub token:

```{r, eval=FALSE}
usethis::create_github_token()
```

This function will open a GitHub webpage. Log in and generate the token. The token will be a series of characters. Save that somewhere, at least temporarily.

Next, run the following and enter the token when asked.

```{r, eval=FALSE}
gitcreds::gitcreds_set()
```

Then, finally, generate the website:

```{r, eval=FALSE}
usethis::use_pkgdown_github_pages()
```

`use_pkgdown_github_pages()` is interactive---you'll have to make a few choices. It performs a number of tasks and then the website is set up. Aside from the website, the most important thing it does is create the `_pkgdown.yml` file. We'll need to edit this later.

Next, we need to convert `README.org` into `README.md`. `pkgdown` doesn't understand Org and requires a README to generate the website's homepage. I used `pandoc`, but I guess you could also convert it by hand.

```
pandoc -s README.org -o README.md
```

Finally, we generate the website:

```{r}
pkgdown::build_site()
```


## See Also

Some documentation that may be useful:

- [`pkgdown` Documentation](https://pkgdown.r-lib.org/index.html).
- [Information on `usethis::use_pkgdown`](https://usethis.r-lib.org/reference/use_pkgdown.html).
- [Managing GitHub Credentials](https://usethis.r-lib.org/articles/git-credentials.html).









