---
title: "Animint2 Quick Start Guide"
---


## Introduction

This quick start guide introduces you to `animint2` in a way that I hope is both brief and friendly. It assumes nothing except some basic knowledge of R and some familiarity with data visualizations and statistics. Some familiarity with `ggplot2` is helpful but unnecessary. After reading this, you'll have a general understanding of how `animint2` works. You'll also know how to generate basic interactive data visualizations, sometimes referred to as "animints." Depending on how sophisticated you want your animints to be, this Quick Start Guide may be all you need.

This guide includes examples of relatively simple animints. `animint2` is also capable of generating larger and more complex data visualizations. For examples, take a look at chapters 8 through 18 of the [animint2 Manual](https://rcdata.nau.edu/genomic-ml/animint2-manual/Ch00-preface.html) or at the [animint gallery](https://rcdata.nau.edu/genomic-ml/animint-gallery/).

The data used in this guide are simulated, which is a fancy way of saying that I forced a computer to make them up. The data are about cats. Here's a glimpse:

```{r, include = FALSE}
# Reproducible random numbers.
set.seed(100)

# Sorting simulated data make the ordered data too perfect. Let's mess it up.
x <- rnorm(n = 100, mean = 5, sd = 1)
y <- runif(n = 100, min = 0, max = 10)
z <- rnorm(n = 100, mean = 5, sd = 1.5) |> sort()
xyz <- rbind(x, y, z) |> c()


# simulate data
Day <- rep(x = 1:30, each = 10)
Cat <- rep(x = c("Clifford", "Junebug", "Muffin", "Teddy Bear", "Diana", "Bello", "Jellybean", "Archibald", "Saturday", "Wilbur"), times = 30)
Cuteness <- xyz |> signif(digits = 2)
Coolness <- xyz |> sort() |> signif(digits = 2)

# concatenate into dataframe
meowtrics <- data.frame(Day, Cat, Coolness, Cuteness)
```

```{r}
head(meowtrics)
```

## Anatomy of a Data Visualization

A data visualization is composed of parts arranged in a certain way. This is often referred to as the grammar of graphics. This quick start guide won't teach you the grammar, and you don't need to know it to get started with `animint2`. What you do need to know is that `animint2`'s (and `ggplot`'s) syntax is modeled on the grammar.

As an example, let's load `animint2` and look at a static graph it produces. It's about perceptions of cat cuteness over time and uses data from 10 different housecats.

```{r}
library(animint2)

cute_1 <- meowtrics |>                                    #1
  ggplot() +                                              #2
  geom_point() +                                          #3
  aes(x = Day, y = Cuteness, group = Cat, color = Cat) +  #4
  labs(title = "Cat Cuteness Ratings Over Time")          #5
cute_1
```

What's happening with the code here?

<dl class="annotated">
	<dt>#1
		<dd>I begin with the dataset, called `meowtrics`.
	<dt>#2
		<dd>I pipe it into `ggplot()`, which tells `animint2` that I'm creating a data visualization.
	<dt>#3
		<dd>I declare that it's a scatterplot using `geom_point()`.
	<dt>#4
		<dd>I draw the axes of the graph with `aes()`. I place the time category in the x-axis and the cuteness ratings in the y-axis (the `x = Day` and `y = Cuteness` arguments, respectively). Using the `group` and `color` arguments, I also group the data together by the cat and color-code them.
	<dt>$5
		<dd>I give the graph a title using `labs()` and its `title` argument.
</dl>

Notice how the data visualization consists of parts---data, axes, points, title, a legend---that are held together using the addition symbol (`+`). All those parts can be manipulated, added, and removed. Different functions and arguments can create different visualizations. Here's how the data would look if it were visualized as a smoothed line graph instead:

```{r}
cute_2 <- meowtrics |>
  ggplot() +
  geom_smooth(se = FALSE) +
  aes(x = Day, y = Cuteness, color = Cat, group = Cat) +
  labs(title = "Cat Cuteness Ratings Over Time")
cute_2
```

The sole difference is that we've replaced the points with smoothed lines. (By default, `geom_smooth()` uses `loess()`.) For stylistic reasons, I've also used the `se = FALSE` argument to remove the confidence internals around the lines.

It's also possible to have both the points and smoothed lines in one graph. Just use both the `geom_point()` abd `geom_smooth()` arguments in the same data visualization:

```{r}
cute_3 <- meowtrics |>
  ggplot() +
  geom_point() + 
  geom_smooth(se = FALSE) +
  aes(x = Day, y = Cuteness, color = Cat, group = Cat) +
  labs(title = "Cat Cuteness Ratings Over Time")
cute_3
```


## How to Make an Animint

The reason why you want to use `animint2` (versus `ggplot2`) is because of the former's support for interactive data visualizations. It's easy to render a static graph interactive. Just use the `animint()` function with the previous plot as the argument:

```{r}
animint(cute_1)
```

This is an animint of our first static data visualization. By clicking on the legend or by using the selection menu, you can control which subjects have their data graphed, as well as how many. This allows you to explore your data without needing to facet all possible cat combinations. For exploratory data analysis, this level of control may be all you need.

In some cases, you may want more control. Say you want to emphasize Archibald and Muffin's cuteness ratings over time. In `animint2`, you would use the `first` argument and specify which cats to present:

```{r}
cute_present <- cute_1 |> 
  animint(first = list(Cat = c("Archibald", "Muffin")))
cute_present
```

This is most useful for situations where you're showcasing or presenting your animint. Pay attention to the syntax: when you use `first`, you must use a `list()`.

Our current animints show three variables: Day, Cuteness, and Cat. What if you want to explore or present a fourth? (Recall that our `meowtrics` dataset has a Coolness variable we haven't looked at yet.) In a static data visualization, this would require multiple graphs or the addition of an unwieldy third dimension.

Luckily, animints are not subject to the same restrictions. In `animint()`, we can use the `showSelected` and `time` arguments to show how the Coolness and Cuteness variables interact day-by-day.

First, let's look at an animint with the `showSelected` variable:

```{r}
associations <- meowtrics |>
  ggplot() +
  geom_point(showSelected = "Day") +
  aes(x = Coolness, y = Cuteness, color = Cat, group = Cat) +
  labs(title = "Associations Between Cuteness and Coolness")
animint(associations)
```

You can use the selection menu to see the different associations between cuteness and coolness ratings per day.

Next, let's see that same animint with the `time` and `duration` option applied:

```{r}
animated_associations <- animint(associations, 
                                 duration = list(Day = 1000),
                                 time = list(variable = "Day", 
                                             ms = 1000))
animated_associations
```

`duration` specifies how quickly the points move from their old location to their new location. The shorter the duration, the quicker the movement. In contrast, `time` specifies how long the points stay in place. In fact, the `duration` argument is optional. If you decide not to use `duration` or set it to 0 milliseconds, the points teleport from one location to another. Click the "Show animation controls" button and play around with it.

Notice how `duration` and `time` also need to take lists. They use milliseconds as their unit of measurement.


## The End

And that's it! You're now a competent `animint2` user. Your invisible and intangible certification of completion will arrive by post within seven business days.

If this quick start guide doesn't satisfy your thirst for animint knowledge, consult the [Animint2 Manual](https://rcdata.nau.edu/genomic-ml/animint2-manual/Ch00-preface.html). Feel free to post any questions to our [GitHub issues](https://github.com/animint/animint2/issues).