---
title: "Animint2 Quick Start Guide"
---


## Introduction

Welcome to the quick start guide. My goal is to introduce you to `animint2` in a way that is both brief and easy to understand. I assume nothing except some basic familiarity with R, data visualizations, and a little bit of statistics. Some familiarity with the `ggplot2` package is helpful but unnecessary.

After reading this, you'll leave with a sense of how `animint2` works. You'll also know how to make interactive data visualizations, sometimes referred to as animints. For basic animints, this quick start guide is all you'll need. To learn how to make more sophisticated animints, take a look at the [animint2 Manual](https://rcdata.nau.edu/genomic-ml/animint2-manual/Ch00-preface.html). You may also want to take a look at the [animint gallery](https://rcdata.nau.edu/genomic-ml/animint-gallery/), which showcases how sophisticated and powerful animints can be.


## Datasets

Some readers may want to follow along interactively. Both `animint2` and R have datasets that you may use. Access the dataset list with `data()`. Alternatively, you may already have a dataset you'd like to use to experiment with `animint2`.

I use a custom dataset in this guide, which I've named `meowtrics`. The data in the dataset are simulated, which is a fancy way of saying that I forced a computer to make them up. The data are about ten different housecats and how they're perceived over time.<sup><a href="#fn-1">i</a></sup> Here's a glimpse:

```{r datasets-make-meowtrics, include = FALSE}
# Reproducible random numbers.
set.seed(100)

# Sorting simulated data make the ordered data too perfect. Let's mess it up.
x <- rnorm(n = 100, mean = 5, sd = 1)
y <- runif(n = 100, min = 0, max = 10)
z <- rnorm(n = 100, mean = 5, sd = 1.5) |> sort()
xyz <- rbind(x, y, z) |> c()


# Simulate data.
Day <- rep(x = 1:30, each = 10)
Cat <- rep(x = c("Clifford", "Junebug", "Muffin", "Teddy Bear", "Diana", "Bello", "Jellybean", "Archibald", "Saturday", "Wilbur"), times = 30)
Kind <- rep(x = c("Domestic Shorthair", "Domestic Longhair", "Cornish Rex", "Domestic Shorthair", "Domestic Shorthair", "Domestic Shorthair", "Siamese", "Maine Coon", "Domestic Longhair", "Domestic Shorthair"), times = 30)
Cuteness <- xyz |> signif(digits = 2)
Coolness <- xyz |> sort() |> signif(digits = 2)

# Concatenate into dataframe.
meowtrics <- data.frame(Day, Cat, Coolness, Cuteness, Kind)
```

```{r datasets-show-meowtrics}
head(meowtrics)
```


## Anatomy of a Data Visualization

Data visualizations are commonplace, and for good reason. Large tables of data are difficult to parse. A good data visualization can illuminate patterns that would have otherwise hard to spot. In contrast, a poor (or deliberately misleading) data visualization can obscure even obvious patterns. How do we design good data visualizations while avoiding bad ones?

We start by understanding what data visualizations are made of and how they're arranged. This arrangement is often called the grammar of graphics.<sup><a href="#fn-2">ii</a></sup> This quick start guide won't teach you the grammar, and you don't need to know it to get started with `animint2`. But you should know that the syntax of `animint2`---that is, the way the code is written---is modelled on the grammar.

Let's see an example. Say I want to visualize how people rate cat cuteness over time. I want the y-axis to depict the cuteness ratings, and the x-axis the days. I also want it to be a scatterplot. What should I do?

First, if I haven't already, I need to install `animint2`:

```{r anatomy-installation, eval=FALSE}
install.packages("animint2")
```

I load `animint2`:

```{r anatomy-load-animint2}
library(animint2)
```

Next, I search for the functions that I'll need and put them together. First, I need `ggplot()`, which is like the blank sheet of paper (or computer screen) the program draws on to make the graph.<sup><a href="#fn-3">iii</a></sup> To make a scatterplot, I'll use either `geom_point()` or `geom_jitter()`. Finally, for the axes, I'll need `aes()`.

Then I name the graph and put it all together:

```{r anatomy-cute-plot, fig.alt = "A scatterplot about cat cuteness ratings over time. The x-axis are the days, and the y-axis the cuteness ratings. There is a lot of variance in the data, but in general cat cuteness ratings rise over time."}
cute_plot <-                    #1
  ggplot() +                    #2
  geom_point(                   #3
    data = meowtrics,           #4
    aes(x = Day, y = Cuteness)) #5
cute_plot                       #6
```

If you're unfamiliar with the syntax, the code can get confusing.<sup><a href="#fn-4">iv</a></sup> Let's go over this code block line by line:

<ol class="annotated">
	<li><span class="explanation">I name the data visualization. Since it's a scatterplot about cat cuteness, I name it `cute_plot`. You can name data visualizations whatever you like, but it's best if you name it something that will make sense to your future self.</span>
	<li><span class="explanation">Next, I call the `ggplot()` function, since I'm making a data visualization.</span>
	<li><span class="explanation">Then I call `geom_point()`, because I'm making a scatterplot.</span>
	<li><span class="explanation">I want to use my `meowtrics` dataset to draw the points in my scatterplot, so I set `data = meowtrics`.</span>
	<li><span class="explanation">`aes()` controls the aesthetics of the data visualization, including axes. I place it inside `geom_point()` and tell my program that the Day and Cuteness variables are on the x- and y-axes, respectively.</span>
	<li><span class="explanation">I repeat the name of my data visualization, which tells the program to display my graph.</span>
</ol>

<!-- Why use HTML instead of Markdown's syntax for ordered lists? I need to control the appearance of this specific ordered list. With Markdown, it's impossible. If you can figure out a way to do that, feel free to change the syntax. :> -->

Look at the plot. We can see that while there's a lot of variance in the data, there seems to be an upward trend in cat cuteness ratings. But this graph could be better. And `animint2` gives us the tools to improve it.

Take a look at this slightly modified data visualization:

```{r anatomy-cute-plot-colored, fig.alt = "A scatterplot titled Cat Cutness Ratings Over Time. It looks exactly the same as the previous graph, except all the data points are differentiated by color. There is also a legend that maps each color to a cat."}
cute_plot_colored <-                             #1
  ggplot() +
  geom_point(
    data = meowtrics,
    aes(x = Day, y = Cuteness,
        group = Cat, color = Cat)) +             #2
  labs(title = "Cat Cuteness Ratings Over Time") #3
cute_plot_2
```

Each cat is differentiated from one another by color, and there's now a legend. Codewise, what's changed?

<ol class="annotated">
  <li><span class="explanation">This data visualization has a different name. If two plots have the same name, one will override the other.</span>
  <li><span class="explanation">I've added new arguments to `aes()`: I've grouped the points in the scatterplot by cat, and I've also differentiated the cats' data points by color.<sup><a href="#fn-5">v</a></sup></span>
  <li><span class="explanation">Using `labs()`, I've given the graph a title.</span>
</ol>

While this new graph communicates more information, it's also somewhat overwhelming. What if we could move some of the data out of the way? Or compare a subset of cat cuteness ratings instead of seeing everything at once? Reader, I have good news.


## Making Animints

The good news is this: `animint2` makes it easy to render a static data visualization interactive. Just use the `animint()` function with the previous plot as the argument:

```{r basics-cute-plot-colored}
animint(cute_plot_colored)
```

This is an animint of our second static data visualization. By clicking on the legend or by using the selection menu, you can control which subjects have their data graphed, as well as how many.<sup><a href="#fn-6">vi</a></sup> This allows you to explore your data without needing to facet all possible cat combinations. For exploratory data analysis, this level of control may be all you need. 

In some cases, you may want more control. Say you want to emphasize Archibald and Muffin's cuteness ratings over time. In `animint2`, you would use the `first` argument and specify which cats to present:

```{r basics-cute-present}
cute_present <- 
  animint(cute_plot_colored, 
          first = list( #1
            Cat = c(    #2
              "Archibald", "Muffin")))
cute_present
```

This is most useful for situations where you're showcasing or presenting your animint. Pay attention to the syntax:

<ol class="annotated">
  <li><span class="explanation">The `first` argument only accepts `list()`s.</span>
  <li><span class="explanation">The specified cats must be in a character vector. That's why they're in a `c()` function.</span>
</ol>


## Using showSelected

<aside class="quick-guide">
<h3 class="ffbug">To Firefox Users</h3>

`animint2` and Firefox [interact strangely](https://github.com/animint/animint2/issues/90). Some animints seem to cause the webpage to scroll up by itself. To get around this, either:

1. Click the "Hide selection menus" button if you see it present; or
2. Click the "Show animation control" button and hit pause.

Thanks! Sorry about the bother.
</aside>

Our current animints use three variables: Day, Cuteness, and Cat. What if you want to explore or present a fourth?<sup><a href="#fn-7">vii</a></sup> In a static data visualization, this would require multiple graphs or the addition of an unwieldy third dimension.

Luckily, animints are not subject to the same restrictions. In `animint()`, we can use the `showSelected` and `time` arguments to show how the Coolness and Cuteness variables interact day-by-day.

First, let's look at an animint with the `showSelected` variable:

```{r showselected-associations}
associations <- 
  ggplot() +
  geom_point(
    data = meowtrics,
    showSelected = "Day",           #1
    aes(x = Coolness, y = Cuteness, 
        color = Cat, group = Cat, 
        key = Cat)) +               #2
  labs(title = "Associations Between Cuteness and Coolness")
animint(associations)
```

<ol class="annotated">
  <li><span class="explanation">`showSelected = "Day"` lets you adjust the day in the selection menu.</span>
  <li><span class="explanation">Most real-world datasets have missing values. `key` accounts for that when transitioning between different days.</span>
</ol>

You can use the selection menu to see the different associations between cuteness and coolness ratings per day.<sup><a href="#fn-8">viii</a></sup>

Next, let's see that same animint with the `time` and `duration` options applied:

```{r showselected-animated-associations}
animated_associations <- 
  animint(associations, 
          duration = list(Day = 1000), #1
          time = list(                 #2
            variable = "Day", 
            ms = 1000))                #3
animated_associations
```

<ol class="annotated">
  <li><span class="explanation">`duration` specifies how quickly the points move from their old location to their new location.<sup><a href="#fn-9">ix</a></sup> The shorter the duration, the quicker the movement. It takes a list and uses milliseconds as its unit of measurement.</span>
  <li><span class="explanation">`time` also takes a list.</span>
  <li><span class="explanation">In contrast to `duration`, `time = list(ms)` specifies how long the points stay in place. It also uses milliseconds as its unit of measurement.</span>
</ol>

You can also click the "Show animation controls" button and manually adjust both the `time` and `duration`. Try it out.


## Using clickSelects

So far, we've been interacting with animints by clicking the legend and using the selection menus and animation controls. What if we could interact with the animint directly? Using `clickSelects`, we can do just that. Let's return to our data visualization depicting cuteness ratings over time, this time as an animint:

```{r clickselects-cute-colored-again}
cute_colored_again <-
  ggplot() +
  geom_point(
    data = meowtrics,
    clickSelects = "Cat", #1
    aes(x = Day, y = Cuteness,
        group = Cat, color = Cat)) +
  labs(title = "Cat Cuteness Ratings Over Time")
animint(cute_colored_again)
```

<ol class="annotated">
  <li><span class="explanation">`clickSelects` takes a variable with quotations. The variable in `clickSelects` and `group` are the same: they're both Cat.</span>
</ol>

Hover over and click on the data points. You'll notice that it has the same effect as clicking on the legend: It removes the data from the animint.

Now, let's interact with a very similar animint:

<!-- this is cute_plot_4 DELETE LATER after updating this -->
```{r clickselects-cute-plot-kind}
cute_plot_kind <-
  ggplot() +
  geom_point(
    data = meowtrics,
    clickSelects = "Kind", #1
    aes(x = Day, y = Cuteness,
        group = Cat, color = Cat)) +
  labs(title = "Cat Cuteness Ratings Over Time")
animint(cute_plot_kind)
```

<ol class="annotated">
  <li><span class="explanation">I've swapped out Cat for Kind. Now the variables in `clickSelects` and `group` are different. This is a useful way of adding another variable to your data visualizations.</span>
</ol>

When you interact with this animint, you'll notice three differences:

1. Hover over the animints. In both, you'll notice a hover box appear. In the first animint, the hover box repeats the cat's name. In the second, the hover box describes what kind of cat they are.
2. In the second animint, there is an additional variable in the selection menu. You can use it to highlight the data points depicting a certain kind of cat.
3. When you click on a data point, it doesn't disappear. Instead, like the selection menu, it highlights data points depicting certain kinds of cats.


## Linked Plots

`animint2` also allows us to link multiple plots together into one animint. When two plots are linked, a change in one plot can cause changes in another.

Let's return to two interactive data visualizations we've already looked at: associations between coolness and cuteness, and cuteness ratings over time.

Here's the first data visualization again, lightly altered:

```{r linked-associations}
associations_again <-
  ggplot() +
  geom_point(
    data = meowtrics,
    showSelected = "Day",
    clickSelects = "Kind",
    aes(x = Coolness, y = Cuteness,
        color = Cat, group = Cat,
        key = Cat))
```

I want to link it to my plot about cuteness ratings over time. That way, I can see how coolness ratings change over time, too. The linking process is a little more involved than usual:

```{r linked-plots}
md <- data.frame(Day = unique(meowtrics$Day))  #1

linked <- animint(associations_again,          #2
                  duration = list(Day = 1000)) #3

linked$overtime <-                             #4
  ggplot() + 
  geom_tallrect(                               #5
    data = md,                                 #6
    aes(xmin = Day-0.5, xmax = Day+0.5),       #7
    clickSelects = "Day",                      #8
    alpha = 0.5                                #9
    ) +                                        #10
  geom_point(
    data = meowtrics,                          #11
    clickSelects = "Cat",
    aes(x = Day, y = Cuteness, 
        group = Cat, color = Cat))
linked                                         #12
```

<ol class="annotated">
  <li><span class="explanation">I'm taking the Day column from `meowtrics` (`meowtrics$Day`), stripping redundant days from it via`unique()`, renaming it, and then making it a `data.frame()`. Now I have a dataframe of all 30 days in `meowtrics`. This is for `geom_tallrect()`, which I'll use later.</span>
  <li><span class="explanation">I take `associations_2` and make it an animint.</span>
  <li><span class="explanation">`duration` controls how long it takes data points to move from one part of the animint to another. As mentioned, it's optional.</span>
  <li><span class="explanation">I take the previously-created animint and attach new animints to it. I name the list of animint `overtime`.</span>
  <li><span class="explanation">This is the `geom_tallrect()` function. It creates a vertical bar that allows you to manipulate time variables by clicking on the plot.</span>
  <li><span class="explanation">`geom_tallrect()` requires a new dataset that contains only the time variable. It's using the dataset that I constructed earlier.</span>
  <li><span class="explanation">`xmin` and `xmax` control how wide the tallrect is, which affects the appearance of the selected time variable. In this case, it controls how much a day takes up on the tallrect.</span>
  <li><span class="explanation">`clickSelects` is necessary here, since we need to be able to select the day from the plot.</span>
  <li><span class="explanation">`alpha` controls the transparency of the tallrect. Lower numbers increase transparency.</span>
  <li><span class="explanation">Using `+`, I attach the tallrect to the plot about cuteness ratings over time.</span>
  <li><span class="explanation">We're back to using `meowtrics` as our dataset.</span>
  <li><span class="explanation">Finally, we intitate the animint.</span>
</ol>

Try it out. When you adjust the day on the bottom plot, the top plot also readjusts. When you click on the legend on the top plot, the bottom plot is also affected. These linked plots can get much complex. For examples, see the [animint gallery](https://rcdata.nau.edu/genomic-ml/animint-gallery/).


## Conclusion

And that's it. You're now a reasonably competent `animint2` user. Feel free to post any questions to our [GitHub issues](https://github.com/animint/animint2/issues), and thanks for reading!


## Footnotes

<ol class="footnotes">
  <li id="fn-1">Fun science fact: Cats are objectively the best animal.
  <li id="fn-2">If you're interested in learning about the grammar and want to get right to the primary source, see Leland Wilkinson's _The Grammar of Graphics_.
  <li id="fn-3">All data visualizations use `ggplot()`.
  <li id="fn-4">Notice also that the `ggplot()` and `geom_point()` functions are held together by the `+` symbol. In other words: You begin with a blank data visualization and then add a scatterplot atop it. All functions in `animint2` are held together with `+`. You'll be using it a lot.
  <li id="fn-5">It's possible to differentiate the data points in a different manner. For example, instead of different colors, I could have used different shapes.
  <li id="fn-6">Play around with the animint. See what you can and can't interact with.
  <li id="fn-7">Recall that our `meowtrics` dataset has a Coolness variable we haven't looked at yet.
  <li id="fn-8">Play around with it and see what you can do.
  <li id="fn-9">Fun fact: The `duration` argument is optional. If you decide not to use `duration` or set it to 0 milliseconds, the points teleport from one location to another.
</ol>







