---
title: "Animint2 Quick Start Guide"
---


## Introduction

Welcome to the quick start guide. My goal is to introduce you to `animint2` in a way that is both brief and easy to understand. I assume nothing except some basic familiarity with R, data visualizations, and a little bit of statistics. Some familiarity with the `ggplot2` package is helpful but unnecessary.

After reading this, you'll leave with a sense of how `animint2` works. You'll also know how to make interactive data visualizations, sometimes referred to as animints. For basic animints, this quick start guide is all you'll need. To learn how to make more sophisticated animints, take a look at the [animint2 Manual](https://rcdata.nau.edu/genomic-ml/animint2-manual/Ch00-preface.html). You may also want to take a look at the [animint gallery](https://rcdata.nau.edu/genomic-ml/animint-gallery/), which showcases how sophisticated and powerful animints can be.


## Datasets

Some readers may want to follow along interactively. Both `animint2` and R have datasets that you may use. Access the dataset list with `data()`. Alternatively, you may already have a dataset you'd like to use to experiment with `animint2`.

I use a custom dataset in this guide, which I've named `meowtrics`. The data in the dataset are simulated, which is a fancy way of saying that I forced a computer to make them up. The data are about ten different housecats and how they're perceived over time.[^cat] Here's a glimpse:

[^cat]: Fun science fact: Cats are objectively the best animal.

```{r datasets-make-meowtrics, include = FALSE}
# Reproducible random numbers.
set.seed(100)

# Sorting simulated data make the ordered data too perfect. Let's mess it up.
x <- rnorm(n = 100, mean = 5, sd = 1)
y <- runif(n = 100, min = 0, max = 10)
z <- rnorm(n = 100, mean = 5, sd = 1.5) |> sort()
xyz <- rbind(x, y, z) |> c()


# Simulate data.
Day <- rep(x = 1:30, each = 10)
Cat <- rep(x = c("Clifford", "Junebug", "Muffin", "Teddy Bear", "Diana", "Bello", "Jellybean", "Archibald", "Saturday", "Wilbur"), times = 30)
Cuteness <- xyz |> signif(digits = 2)
Coolness <- xyz |> sort() |> signif(digits = 2)

# Concatenate into dataframe.
meowtrics <- data.frame(Day, Cat, Coolness, Cuteness)
```

```{r datasets-show-meowtrics}
head(meowtrics)
```


## Anatomy of a Data Visualization

Data visualizations are commonplace, and for good reason. Large tables of data are difficult to parse. A good data visualization can illuminate patterns that would have otherwise hard to spot. In contrast, a poor (or deliberately misleading) data visualization can obscure even obvious patterns. How do we design good data visualizations while avoiding bad ones?

We start by understanding what data visualizations are made of and how they're arranged. This arrangement is often called the grammar of graphics. This quick start guide won't teach you the grammar, and you don't need to know it to get started with `animint2`.[^1] But you should know that the syntax of `animint2`---that is, the way the code is written---is modelled on the grammar.

[^1]: If you're interested in learning about the grammar and want to get right to the source, see Leland Wilkinson's _The Grammar of Graphics_.]

Let's see an example. Say I want to visualize how people rate cat cuteness over time. I want the y-axis to depict the cuteness ratings, and the x-axis the days. I also want it to be a scatterplot. What should I do?

First, if I haven't already, I need to install `animint2`:

```{r anatomy-installation, eval=FALSE}
install.packages("animint2")
```

I load `animint2`:

```{r anatomy-load-animint2}
library(animint2)
```

Next, I search for the functions that I'll need and put them together. First, I need `ggplot()`, which is the blank sheet of paper (or computer screen) the code draws on to make the graph.[^2] I'm making a scatterplot, so I'll want something like `geom_point()` or `geom_jitter()`. Finally, for the axes, I'll need `aes()`.

[^2]: All data visualizations use `ggplot()`.

Then I name the graph and put it all together:

```{r anatomy-cute-plot, fig.alt = "A scatterplot about cat cuteness ratings over time. The x-axis are the days, and the y-axis the cuteness ratings. There is a lot of variance in the data, but in general cat cuteness ratings rise over time."}
cute_plot <-                    #1
  ggplot() +                    #2
  geom_point(                   #3
    data = meowtrics,           #4
    aes(x = Day, y = Cuteness)) #5
cute_plot                       #6
```

If you're unfamiliar with the syntax, the code can get confusing.[^3] Let's go over this code block line by line:

[^3]: Notice also that the `ggplot()` and `geom_point()` functions are held together by the `+` symbol. In other words: You begin with a blank data visualization and then add a scatterplot atop it. All functions in `animint2` are held together with `+`. You'll be using it a lot.

<ol class="annotated">
	<li><span class="explanation">I name the data visualization. Since it's a scatterplot about cat cuteness, I name it `cute_plot`. You can name data visualizations whatever you like, but it's best if you name it something that will make sense to other programmers. Trust me---your future self will thank you.</span>
	<li><span class="explanation">Next, I call the `ggplot()` function, since I'm making a data visualization.</span>
	<li><span class="explanation">Then I call `geom_point()`, because I'm making a scatterplot.</span>
	<li><span class="explanation">I want to use my `meowtrics` dataset to draw the points in my scatterplot, so I set `data = meowtrics`.</span>
	<li><span class="explanation">`aes()` controls the aesthetics of the data visualization, including axes. I place it inside `geom_point()` and tell my ptogram that the Day and Cuteness variables are on the x- and y-axes, respectively.</span>
	<li><span class="explanation">I repeat the name of my data visualization, which tells the program to display my graph.</span>

<!-- Why use HTML instead of Markdown's syntax for ordered lists? I need to control the appearance of this specific ordered list. With Markdown, it's impossible. If you can figure out a way to do that, feel free to change the syntax. :> -->

Our code has produced a useful graph. We can see that while there's a lot of variance in the data, there seems to be an upward trend in cat cuteness ratings. But this graph could be better. And `animint2` gives us the tools to improve it.

Take a look at this slightly modified data visualization:

```{r anatomy-cute-plot-2, fig.alt = "A scatterplot titled Cat Cutness Ratings Over Time. It looks exactly the same as the previous graph, except all the data points are differentiated by color. There is also a legend that maps each color to a cat."}
cute_plot_2 <-                                   #1
  ggplot() +
  geom_point(
    data = meowtrics,
    aes(x = Day, y = Cuteness,
        group = Cat, color = Cat)) +             #2
  labs(title = "Cat Cuteness Ratings Over Time") #3
cute_plot_2
```

Each cat is differentiated from one another by color. I've also added a color, and there's now a legend. Codewise, what's changed?

<ol class="annotated">
  <li><span class="explanation">This data visualization has a different name. Make sure all your names are different---if they aren't, later visualizations will override the former ones.</span>
  <li><span class="explanation">I've added new arguments to `aes()`: I've grouped the points in the scatterplot by cat. I've also differentiated the cats' data points by color.[^4]</span>
  <li><span class="explanation">Using `labs()`, I've given the graph a title.</span>
</ol>

<!-- See the comment above this one. :> -->

[^4]: It's possible to diffentiate the data points in a different manner. For example, instead of different colors, I could have used different shapes.

While this new graph adds more information, it's difficult to use some of it. There's just a lot more being communicated. What if we could move some of the data out of the way? Or compare a subset of cat cuteness ratings instead of seeing everything at once? Reader, I have good news.


## Making Animints

The good news is this: `animint2` makes it easy to render a static data visualization interactive. Just use the `animint()` function with the previous plot as the argument:

```{r animints-cute-plot-2}
animint(cute_plot_2)
```

This is an animint of our second static data visualization. By clicking on the legend or by using the selection menu, you can control which subjects have their data graphed, as well as how many.[^5] This allows you to explore your data without needing to facet all possible cat combinations. For exploratory data analysis, this level of control may be all you need.

[^5]: Play around with the animint. See what you can and can't interact with.

In some cases, you may want more control. Say you want to emphasize Archibald and Muffin's cuteness ratings over time. In `animint2`, you would use the `first` argument and specify which cats to present:

```{r animints-cute-present}
cute_present <- 
  animint(cute_plot_2, 
          first = list( #1
            Cat = c(    #2
              "Archibald", "Muffin")))
cute_present
```

This is most useful for situations where you're showcasing or presenting your animint. Pay attention to the syntax:

<ol class="annotated">
  <li><span class="explanation">The `first` argument only accepts `list()`s.</span>
  <li><span class="explanation">The specific cats must also be in a list. That's why they're in a `c()` function.</span>
</ol>

<!-- See the comment above this one. :> -->


### Using showSelected

Our current animints use three variables: Day, Cuteness, and Cat. What if you want to explore or present a fourth?[^6] In a static data visualization, this would require multiple graphs or the addition of an unwieldy third dimension.

[^6]: Recall that our `meowtrics` dataset has a Coolness variable we haven't looked at yet.

Luckily, animints are not subject to the same restrictions. In `animint()`, we can use the `showSelected` and `time` arguments to show how the Coolness and Cuteness variables interact day-by-day.

First, let's look at an animint with the `showSelected` variable:

```{r showselected-associations}
associations <- 
  ggplot() +
  geom_point(
    data = meowtrics,
    showSelected = "Day",           #1
    aes(x = Coolness, y = Cuteness, 
        color = Cat, group = Cat, 
        key = Cat)) +               #2
  labs(title = "Associations Between Cuteness and Coolness")
animint(associations)
```

<ol class="annotated">
  <li><span class="explanation">`showSelected` lets you adjust the day in the selection menu. It takes a variable with quotations.</span>
  <li><span class="explanation">Most real-world datasets have missing values. Here, `key` accounts for that when transitioning between different days. It takes a variable without quotations.</span>
</ol>

<!-- See the comment above this one. :> -->

You can use the selection menu to see the different associations between cuteness and coolness ratings per day.[^7]

[^7]: Play around with it and see what you can do.

Next, let's see that same animint with the `time` and `duration` option applied:

```{r showselected-animated-associations}
animated_associations <- 
  animint(associations, 
          duration = list(Day = 1000), #1
          time = list(                 #2
            variable = "Day", 
            ms = 1000))                #3
animated_associations
```

<ol class="annotated">
  <li><span class="explanation">`duration` specifies how quickly the points move from their old location to their new location.[^8] The shorter the duration, the quicker the movement. It takes a list and uses milliseconds as its unit of measurement.</span>
  <li><span class="explanation">`time` also takes a list.</span>
  <li><span class="explanation">In contrast to `duration`, `time = list(ms)` specifies how long the points stay in place. It also uses milliseconds as its unit of measurement.</span>
  <li><span class="explanation"></span>
</ol>

<!-- See the comment above this one. :> -->

[^8]: Fun fact: The `duration` argument is optional. If you decide not to use `duration` or set it to 0 milliseconds, the points teleport from one location to another.

You can also click the "Show animation controls" button and manually adjust both the `time` and `duration`. Try it out.


### Using clickSelects

empty for now


## Goodbye

And that's it. You're now a reasonably competent `animint2` user. Your certificate of completion will arrive by post within four to five business days.[^coc]

[^coc]: The certificate of completion is both invisible and intangible. Some philosophers posit that this certificate may not exist at all.

Feel free to post any questions to our [GitHub issues](https://github.com/animint/animint2/issues). Thanks for reading.