---
title: "How was this Website Set Up?"
date: First written 2023-06-28. Last revised 2023-07-20.
---

This is an internal document intended for `animint2` contributors. It documents how I set up the reference website, as well as the code I ran to do so. It may be useful for debugging. It should also be useful if, for whatever reason, you need to set up the website again.

I've tried to make this document and the website-generating procedures clear and near-algorithmic as I reasonably can. Inevitably, some of this may seem condescending. Sorry if it is.


## Get Ready

First, set your working directory to the `animint2` repository. If you're using an command line terminal, type in `cd whatever/your/path/is/animint/animint2`. If you're using RStudio, you can instead use the <kbd>CTRL</kbd><kbd>SHIFT</kbd><kbd>H</kbd> keyboard shortcut and select the `animint2` repository from there.

Next, create a `README.md` document if it doesn't exist. Toby already has a `README.org` document, but `pkgdown` doesn't understand Org. I used `pandoc`, a document converter, but you can also do it by hand:

```
pandoc -s README.org -o README.md
```

Then generate a GitHub token. It's like a more secure password and is necessary to access GitHub's API, which we need to set up the website:

```{r, eval=FALSE}
usethis::create_github_token()
```

This function will open a GitHub webpage. Log in and generate the token. The token will be a series of characters. Save that somewhere, at least temporarily.

Run the following and enter the token when asked:

```{r, eval=FALSE}
gitcreds::gitcreds_set()
```

Finally, configure the repository to use `pkgdown`:

```{r, eval=FALSE}
usethis::use_pkgdown()
```

That's it. Now we can build the site.


## Build the Site

This section assumes that everything in the previous section has already happened, but that you need to build the website again.

We build the site, then get it ready for GitHub Pages:

```{r, eval=FALSE}
pkgdown::build_site()
usethis::use_pkgdown_github_pages()
```

`pkgdown::build_site()` is the most important, and generates all the HTML files for our website. `usethis::use_pkgdown_github_pages()` creates or modifies the `_pkgdown.yml` file, which influences how the site looks. If the `_pkgdown.yml` file is already how you like it, just say no when `usethis` offers to modify it.

And you're done with site building.


## Modify the Site

This section assumes that everything in the previous two sections has already happened. You're reading the section to figure out how to make changes to the website. By default, the home page is just a mirror image of what's on the `animint2` repository, and the references are in alphabetical order. We can make it more useful.




## Make the Site Appear on GitHub Pages

x


## See Also

Some documentation that may be useful:

- [`pkgdown` Documentation](https://pkgdown.r-lib.org/index.html).
- [Information on `usethis::use_pkgdown`](https://usethis.r-lib.org/reference/use_pkgdown.html).
- [Managing GitHub Credentials](https://usethis.r-lib.org/articles/git-credentials.html).
- [pandoc Home Page](https://pandoc.org/).









