---
title: "How was this Website Set Up?"
date: First written 2023-06-28. Last revised 2023-07-21.
---

This is an internal document intended for `animint2` contributors. It documents how I set up the reference website, as well as the code I ran to do so. It may be useful for debugging. It should also be useful if, for whatever reason, you need to set up the website again.

I've tried to make this document and the website-generating procedures clear and near-algorithmic as I reasonably can. Inevitably, some of this may seem condescending. Sorry if it is.


## Get Ready

First, set your working directory to the `animint2` repository. If you're using an command line terminal, type in `cd whatever/your/path/is/animint/animint2`. If you're using RStudio, you can instead use the <kbd>CTRL</kbd><kbd>SHIFT</kbd><kbd>H</kbd> keyboard shortcut and select the `animint2` repository from there.

A `README.md` document should already exist. If it doesn't, create one. Remember to use [GitHub-flavored Markdown](https://github.github.com/gfm/). The `README` should have, at minimum:

1. A section that briefly summarizes what the package is about,
2. A section that teaches the reader how to install both the CRAN and development versions of the package, and
3. A section that briefly demonstrates how to do basic things with the package or links to another page that does.

In my `README`, I included About section with a GIF of someone using `animint2`, an Installation section, a Use section that linked to a quick start guide, a Similar Packages section about other packages that created interactive data visualizations, and a Problems? section asking users to report bugs if they find any.

Next, generate a GitHub token. It's like a more secure password. We use it to access GitHub's API, which we need to set up the website:

```{r, eval = FALSE}
usethis::create_github_token()
```

This function will open a GitHub webpage. Log in and generate the token. The token will be a series of characters. Save that somewhere, at least temporarily.

Run the following and enter the token when asked:

```{r, eval = FALSE}
gitcreds::gitcreds_set()
```

Finally, configure the repository to use `pkgdown`:

```{r, eval = FALSE}
usethis::use_pkgdown()
```

That's it. Now we can build the site.


## Build the Site

We can build the whole site with just a single line of code:

```{r, eval = FALSE}
pkgdown::build_site()
```

This generates HTML files from our `.Rd` files.

With another line of code, we create and edit the `_pkgdown.yml` file, which influences how the site looks:

```{r, eval = FALSE}
usethis::use_pkgdown_github_pages()
```

If there's already a `_pkgdown.yml` file properly set up, just say no when `usethis` offers to modify it.

And you're done with site building.


## Modify the Site

By default, the home page is just a mirror image of what's on the `animint2` repository, and the references are in alphabetical order. Not great. We can make it better.

### Rearranging the Reference Pages

You can [rearrange the reference pages](https://pkgdown.r-lib.org/reference/build_reference.html) by adding a `reference` section to the `_pkgdown.yml`. I've linked to the `pkgdown` documentation, which describes the syntax in detail. It's gonna take a while to build, in part due to the sheer number of functions `animint2` has. Two tricks that helped me:

- Use `data()` to pull up all the datasets that `animint2` has. (Thirty-three!) Then just copy/paste the list into a Dataset section in `_pkgdown.yaml`.
- Use [`ggplot2`'s YAML file](https://github.com/tidyverse/ggplot2/blob/main/_pkgdown.yml) as a reference. `animint2` is a fork of `ggplot2`, so there's lots of similarities. Don't mindlessly copy/paste, though.

### Change the Look

empty


## Make the Site Appear on GitHub Pages

empty


## See Also

Some documentation that may be useful:

- [Managing GitHub Credentials](https://usethis.r-lib.org/articles/git-credentials.html).
- [`pkgdown` Documentation](https://pkgdown.r-lib.org/index.html).
- [Information on `usethis::use_pkgdown`](https://usethis.r-lib.org/reference/use_pkgdown.html).
- [GitHub-flavored Markdown Specification](https://github.github.com/gfm/)/